Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-11-07 pkg/glmertree/paper/treatment_interactions.Rnw
Several changes in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1013 255f4249-2263-4417-b1ea-8498779e27a7

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure. Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM trees and compare it with that of trees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy (.94 and .88 on average, respectively) and Type-I error rates (4\% and 33\%, respectively). Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed equal predictive accuracy (.60 on average), but better accuracy in detecting interactions. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising algorithm for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model trees, or GLMM trees.

@@

We created a dataset representing observations on 150 participants in a randomized clinical trial. Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the posttreatment total score on a depression inventory. For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).

@@

Participants were part of one of ten clusters, each with a different value for the random intercept. Data were generated such that covariates and cluster-specific intercepts were uncorrelated. 43\% of variance in posttreatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.

@@

For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function\footnote{An overview of notation used is provided in the appendix.}:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper. Also, for the continuous response variable in the motivating data set, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accomodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
For a simple random intercept $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$. For simplicity, we just employ such a cluster-specific intercept here but further random effects can be easily included in $z_i$. Furthermore, within the GLMM, it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

@@

For estimating GLMM trees, we take this approach two steps further: (1)~Instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree. This allows not only for detection of differences in intercepts/levels across terminal nodes, but also for detection of differences in slopes/regression coefficients such as treatment effects. (2)~By using generalized linear (mixed) models, the response may also be a binary or count variable instead of a continuous variable. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

@@

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented. As already mentioned above, by taking into account the clustering of observations, the true treatment subgroups have been recovered, and the spurious split involving the anxiety variable no longer appears in the tree.

\section{Simulation-based evaluation: Method}

To assess the performance of GLMM tree, we carried out three simulation studies. In Study I, we compared the accuracy of GLMM tree with that of GLM tree in datasets with treatment-subgroup interactions. In Study II, we compared the accuracy of GLMM tree with that of GLM tree in datasets without treatment-subgroup interactions, allowing for assessment of the Type-I error rate. In Study III, we compared the accuracy of GLMM tree with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

@@

R \cite{R} was used for the generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| is available for other response distributions. For estimation of GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.
%
For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge). This package provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.

In all simulations, the significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing. The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight.

\subsection{Simulation design}

\readme{Z: My impression is that the description of the data-generating process is too far ahead of the simulation itself. I would recommend to first explain what is the same in all three simulation studies. And then what is specific could maybe be included at the beginning of each of the individual studies so that readers have it in mind when looking at the results. So in this section here I would just briefly explain the idea and defer all details until later.}

\subsubsection{Treatment-subgroup interactions} 
For generating datasets with treatment-subgroup interactions, we used one of the treatment-subgroup interaction designs from \citeA{DussyMech14}, which is depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows two subgroups with mean differences in treatment outcomes, and two subgroups without mean differences in treatment outcomes. The four subgroups are characterized by values on the (true) partitioning variables $U_2$, and $U_1$ or $U_5$. The other potential partitioning variables ($U_3$, $U_4$, $U_6$ through $U_{15}$) are noise variables.

@@

\subsubsection{Continuous interactions}

The subgroups depicted in Figure~\ref{fig:modelC} are piecewise functions of the partitioning variables, therefore these interactions can be referred to as piecewise interactions. However, interactions may also consist of continuous functions of the partitioning variables. For comparing the accuracy of GLMM tree with that of GLMMs with pre-specified interaction terms, we generated datasets in which interactions were varied to be purely piecewise, purely continuous, or a combination of both. To generate purely continuous interactions, the outcome variable $y$ was generated as the sum of:

\begin{enumerate}
	\item a main effect of $U_2$ (coefficient = 1.5) 
	\item an interaction between $U_1$ and $U_2$ with a negative effect (coefficient = $-.25$), and an interaction between $U_2$ and $U_5$ with a positive effect (coefficient = $.25$)
	\item an interaction between $U_1$, $U_2$ and treatment with a negative effect (coefficient = $-.25$), and an interaction between $U_2$, $U_5$ and the treatment indicator with a positive effect (coefficient = $.25$)
\end{enumerate}

\readme{Z: Also include the formula for the linear predictor in this setup so that this can be easily matched with the GLMM estimated.}

Note that the direction (positive or negative) and type (main or interaction) of these effects correspond to those in Figure~\ref{fig:modelC}. All interactions were created by using centered $U_k$ variables, calculated by subtracting the variable means.

\subsubsection{Data-generating parameters} 
In generating datasets, we varied seven parameters of the data-generating process:

\begin{enumerate} 
\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$, $K=15$ (where only $U_1$, $U_2$ and $U_5$ are true partitioning variables).
\item Two levels of intercorrelations between the covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Three levels for the intercorrelations between $b$ and one of the $U_k$ variables: $b$ and $U_k$ uncorrelated, $b$ correlated with a true partitioning variable (i.e., $U_2$, $U_1$, or $U_5$, introducing a correlation of $\approx 0.42$), $b$ correlated with a non-partitioning covariate (i.e., $U_3$ or $U_4$, introducing a correlation of $\approx 0.42$)\footnote{Note that, when $\sigma_b = 0$, the correlation between $b$ and one of the $U_k$ variables is 0, by definition. However, datasets were created for this condition, to allow for a full factorial design of the simulation study; in reality, $b$ and $U$ are uncorrelated in these instances.}. 
\item Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes. In datasets with treatment-subgroup interactions, the treatment effect difference varied only in nodes 4 and 7, with levels $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

\subsubsection{Variable distributions}
Following the approach of \citeA{DussyMech14}, all covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. The means for all other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ have the same standard deviation: ${\sigma_U}_k = 10$. Correlations between $U$ variables vary according to the third facet of the data-generating design described above.

To generate the cluster-specific intercepts $b_m$, we partitioned the sample into equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design. For each cluster, a single value $b_m$ was drawn from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design. When $b$ was correlated with one of the potential partitioning variables, this variable was randomly selected.

To generate node-specific fixed effects, we partitioned the sample according to the terminal nodes of the tree in Figure 4.3. In combination with the seventh facet of the simulation design, this determines the values of $\beta_j$. For every observation, we generated a binomial variable (with probability .5) as an indicator for treatment type. Random errors $\epsilon$ were drawn from a normal distribution with $\mu_{\epsilon} = 0$ and $\sigma_{\epsilon} = 5$.

Finally, the response variable was calculated as the sum of the (node-specific) fixed effects, the random intercept and the error term: $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.

@@

Firstly, the total number of nodes in estimated GLM and GLMM tree were calculated to assess the extent to which the algorithms may detect spurious subgroups. Also, the number of nodes allow for assessing Type-I (the extent to which datasets are erroneously partitioned) and Type-II (the extent to which datasets are erroneously not partitioned) error.

Secondly, in datasets with treatment-subgroup interactions, we assessed the accuracy of estimated models. For GLM and GLMM trees, an accurately recovered tree was defined as a tree with (1) the true tree size (i.e., tree size $= 7$ in datasets with treatment-subgroup interactions), (2) the first split in the tree involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$. Note that the allowance of $\pm 5$ equals an allowance of plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).

Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences ($\beta_{j1}$ in Figure~\ref{fig:modelC}) in each dataset. Consequently, predictive accuracy was only assessed in datasets with treatment-subgroup interactions, as the true treatment differences have a constant value in datasets without treatment-subgroup interactions. To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets. Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from `new' clusters.

\section{Simulation-based evaluation: Results}

\subsection{Study I: Detection of Treatment-Subgroup Interactions}

For each cell of the data-generating design described above, 50 datasets were generated, resulting in $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2$ = 32,400 training datasets with treatment-subgroup interactions. In these datasets, the true tree size was 7 (4 terminal nodes and 3 inner nodes; Figure~\ref{fig:modelC}).

@@

\begin{figure}[t!]
\centering
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T))
)
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent levels of sample size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

@@

GLMM tree yielded trees with an average size of 7.15 nodes ($\mathrm{SD}=0.61$), whereas GLM tree yielded an average tree size of 8.15 nodes ($\mathrm{SD}=2.05$), indicating that GLM trees involved more spurious splits than GLMM trees, on average. The estimated probability that a dataset was erroneously not partitioned (the Type-II error) was 0 for both algorithms.

To find the most important predictors of tree size, we performed an ANOVA with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors of tree size were algorithm type, sample size, variance of the random intercept, and the correlation between partitioning variables. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treesize_interact}).

Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions. On average, tree size increases slightly with sample size for GLMM tree, and dramatically for GLM tree. In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size. However, clear differences in tree size were observed when $\sigma_b > 0$. When sample size is small and random intercepts are not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average. When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized. However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially with larger sample sizes (i.e., 500 or 1000) and larger $\sigma_b$ values (i.e., 10). This effect was somewhat stronger when the random intercept was correlated with a covariate that did not appear in the true tree.

@@

However, further splits were more accurately recovered by GLMM tree, which accurately recovered the partition in 90.40\% of datasets, and GLM tree in only 61.44\% of datasets. To find the most important predictors of tree accuracy, we fitted a GLM with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and the level of dependence between the partitioning variables and the random intercept. These effects were further assessed by means of a graphical display (Figure~\ref{fig:xyplot_treeaccuracy}).

>>>>

\section{Abstract}
Identification of subgroups of patients for which treatment A is more effective than treatment B, and vice versa, is of key importance to the development of personalized medicine. Tree-based algorithms are helpful tools for the  detection of such interactions, but none of the available tree-based algorithms allow for taking into account a clustered or nested structure. Therefore, we propose the generalized linear mixed-effects model trees (GLMM tree) algorithm, which allows for the detection of treatment-subgroup interactions, while accounting for the clustered structure of a dataset. The algorithm uses model-based recursive partitioning (MOB) to detect treatment-subgroup interactions, and a GLMM for the estimation of random-effects parameters. In a simulation study, we evaluate the performance of GLMM trees and compare it with that of trees without random effects, and GLMMs with pre-specified interactions. GLMM tree was found to accurately recover treatment-subgroup interactions in 90\% of the simulated datasets, whereas trees without random effects only did so in 61\% of datasets. GLMM tree also outperformed trees without random effects in terms of predictive accuracy and Type-I error rates. Furthermore, compared to GLMMs with pre-specified interaction effects, GLMM tree showed better predictive accuracy. We illustrate the application of GLMM trees on an individual patient-level data meta-analysis on treatments for depression. We conclude that GLMM tree is a promising exploratory tool for the detection of treatment-subgroup interactions in clustered datasets.\\
\\
\textit{Keywords}: model-based recursive partitioning, treatment-subgroup interactions, random effects, generalized linear mixed-effects model, classification and regression trees

@@

In many instances, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. For tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection allow for taking into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models: generalized linear mixed-effects model tree, or GLMM tree.

@@

We created a dataset representing observations on 150 participants in a randomized clinical trial. Every participant was randomly assigned to Treatment~1 or Treatment~2, and has a value for the response variable, with which the effect of treatment is assessed: the pos-ttreatment total score on a depression inventory. For all participants, three covariate values are available: duration of depressive symptoms prior to treatment in months (duration, range 0--15); age in years (age, range 18--75); anxiety inventory total score (anxiety, range 3--18).

@@

Participants were part of one of ten clusters, each with a different value for the random intercept. Data were generated such that covariates and cluster-specific intercepts were uncorrelated. 43\% of variance in post-treatment depression scores was due to treatment-subgroup interactions, while 8\% of variance was due to cluster-specific variation.

@@

For example, let us take a global GLM to estimate the overall treatment effect in the motivating dataset. The expectation $\mu_i$ of outcome $y_i$ given the treatment regressors $x_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
	\label{eq:expected_value}
	E[y_i | x_i] & = & \mu_i, \\
	\label{eq:fixedeffects}
	g(\mu_{i}) & = & x_{i}^{\top}\beta,
\end{eqnarray}
%
where $x_{i}^{\top}\beta$ is the linear predictor for observation $i$ and $g$ is the link function. $\beta$ is a vector of fixed-effects regression coefficients, the first element representing the intercept, corresponding to the mean value of the linear predictor in the first treatment group, and the second element representing the slope, which is the mean difference in the linear predictor between the first and second treatment groups. Thus, for simplicity we assume $x_{i}$ and $\beta$ to have length 2 in the current paper. Also, for the continuous response variable in the motivating data set, we employ a Gaussian distribution with identity link and denote the error by $\epsilon_i = y_i - \mu_i$ with variance $\sigma_{\epsilon}^2$. However, the model can easily accommodate additional treatment conditions and covariates, and binary or count outcome variables.

@@

For datasets containing observations from multiple clusters (e.g., trials or research centers), application of a GLMM would be more appropriate. The model in Equation~\ref{eq:fixedeffects} is then extended to include cluster-specific, or random effects:
%
\begin{equation}
	\label{eq:mixedeffects}
	g(\mu_{i}) = x_{i}^{\top}\beta + z_{i}^{\top}b
\end{equation}
%
For a simple random intercept $z_{i}$ is a unit vector of length $M$, of which the $m$-th element takes a value of 1, and all other elements take a value of 0; $m$ ($m=1,\dots,M$) denotes the cluster which observation $i$ is part of. Further, $b$ is a random vector of length $M$, each $m$-th element representing the random intercept for cluster $m$. For simplicity we only employ a cluster-specific intercept here, but further random effects can be easily included in $z_i$. Furthermore, within the GLMM it is assumed that $b$ is normally distributed, with mean zero and variance $\sigma^{2}_{b}$. The parameters of the GLMM can be estimated with, for example, maximum likelihood (ML) and restricted ML (REML).

@@

For estimating GLMM trees, we take this approach two steps further: (1)~Instead of a CART tree with constant fits to estimate the fixed-effects part of the GLMM, we use a GLM tree. This allows not only for detection of differences in intercepts across terminal nodes, but also for detection of differences in slopes such as treatment effects. (2)~By using generalized linear (mixed) models, the response may also be a binary or count variable instead of a continuous variable. The GLMM tree algorithm takes the following steps to estimate the model in Equation~\ref{eq:glimmertree}:
\begin{description}
  \item[Step 0:] Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.

@@

In Figure~\ref{fig:example_glimmertree}, the result of applying the GLMM tree algorithm to the motivating dataset is presented. By taking into account the clustering of observations, the spurious split involving the anxiety variable no longer appears in the tree and the true treatment subgroups have been recovered.

\section{Simulation-based evaluation}

To assess the performance of GLMM tree, we carried out three simulation studies: In Studies I and II, we compared the performance of GLMM tree with that of GLM tree. In Study I, we assessed the accuracy of both tree-based algorithms in datasets with treatment-subgroup interactions. In Study II, we assessed the Type-I error of both tree-based algorithms in datasets without treatment-subgroup interactions. In Study III, we compared the performance of GLMM tree in detecting treatment interactions with that of a more classical approach of interaction detection: a GLMM with pre-specified interactions.

\subsection{General simulation design}

In all simulation studies, we varied the following data-generating parameters:

\begin{enumerate} 
\item Three levels for sample size: $N=200$, $N=500$, $N=1000$.
\item Two levels for the number of potential partitioning covariates $U_1$ through $U_K$: $K=5$ and $K=15$.
\item Two levels for the intercorrelation between the potential partitioning covariates $U_1$ through $U_K$: $\rho_{U_k,U_{k'}}=0.0$, $\rho_{U_k,U_{k'}}=0.3$.
\item Three levels for the number of clusters: $M=5$, $M=10$, $M=25$.
\item Three levels for the population standard deviation of the normal distribution from which the cluster specific intercepts were drawn: $\sigma_{b}=0$, $\sigma_{b}=5$, $\sigma_{b}=10$.
\item Two levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the $U_k$ covariates.  
\end{enumerate}

In addition, each simulation study required a different design for main and interaction effects, which will be described in their respective sections below.

Following the approach of \citeA{DussyMech14}, all partitioning covariates $U_1$ through $U_{K}$ were drawn from a multivariate normal distribution with means ${\mu_U}_1 = 10$, ${\mu_U}_2 = 30$, ${\mu_U}_4 = -40$, and ${\mu_U}_5 = 70$. Means for the other covariates (i.e., ${\mu_U}_3$, and ${\mu_U}_6$ through ${\mu_U}_{15}$) were drawn from a discrete uniform distribution on the interval $[-70,70]$. All covariates $U_1$ through $U_{15}$ had the same standard deviation: ${\sigma_U}_k = 10$. Correlations between $U$ variables vary according to the third facet of the data-generating design described above.

To generate the cluster-specific intercepts $b_m$, we partitioned the sample into $M$ equally-sized clusters, conditional on one of the variables $U_1$ through $U_5$, producing the correlations in the sixth facet of the simulation design. For each cluster, a single value $b_m$ was drawn from a normal distribution with mean 0 and the value of $\sigma_{b}$ given by the fifth facet of the simulation design. When $b$ was correlated with one of the potential partitioning variables, this variable was randomly selected.

For every observation, we generated a binomial variable (with probability .5) as an indicator for treatment type. Random errors $\epsilon$ were drawn from a normal distribution with $\mu_{\epsilon} = 0$ and $\sigma_{\epsilon} = 5$. The value of the outcome variable $y_i$ was calculated as the sum of the random intercept, (node-specific) fixed effects and the random error term.

@@

R \cite{R} was used for the generation and analysis of all datasets. The \verb|partykit| package (version 1.0-2; \citeNP{HothyZeil15}) was employed for estimating GLM trees, using the \verb|lmtree| function for normal linear regressions and \verb|glmtree| for other response distributions. For estimating GLMMs the \verb|lmer| function (or \verb|glmer| function, respectively) from the \verb|lme4| package (version 1.1-7; \citeNP{BateyMeac12}) was employed, using restricted maximum likelihood (REML) estimation.

For estimation of GLMM trees the former two packages were combined in a new package \verb|glmertree| (version 0.1-0; \citeNP{FokkyZeil15}; available from R-Forge). This package provides functions \verb|lmertree| and \verb|glmertree| that iterate between estimation of the \verb|lmtree|/\verb|glmtree| model and the \verb|lmer|/\verb|glmer| model.

The significance level $\alpha$ for the parameter instability tests in the trees was set to .05, with a Bonferroni correction applied for multiple testing, in all simulations. The minimum number of observations per node in the tree was set to 20 and the maximum tree depth was set to four, thus limiting the number of potential subgroups to eight.

In Study III, the \verb|lmerTest| package (version 2.0-32; \citeNP{KuznyBroc16}) was used to assess statistical significance of fixed-effects predictors in GLMMs. The \verb|lmerTest| package calculates effective degrees of freedom and $p$-values based on Satterthwaite approximations.

\subsection{Study I: Method}

@@

\subsubsection{Treatment-subgroup interaction design}

For generating datasets with treatment-subgroup (or piecewise) interactions, we used a design from \citeA{DussyMech14} which is depicted in Figure~\ref{fig:modelC}. Figure~\ref{fig:modelC} shows four subgroups, characterized by values of the (true) partitioning variables $U_2$, and $U_1$ or $U_5$. Two of the subgroups have mean differences in treatment outcome, indicated by a non-zero value of $\beta_{j1}$, and two subgroups do not have mean differences in treatment outcome, indicated by a $\beta_{j1}$ value of 0.

In this simulation study, some of the potential partitioning covariates were true partitioning covariates ($U_1$, $U_2$ and $U_5$), whereas others were not (e.g., $U_3$ and $U_4$). Therefore, an extra level was added to the sixth facet of the \textit{General simulation design} described above. Also, to assess the effect of the size of the treatment-effect differences in nodes 3 and 7, an additional facet was added:

\begin{enumerate}
\item[6.] Three levels for the intercorrelation between $b$ and one of the $U_k$ variables: $b$ and all $U_k$ covariates uncorrelated, $b$ correlated with one of the true partitioning covariates ($U_1$, $U_2$ or $U_5$), $b$ correlated with one of the noise variables ($U_3$ or $U_4$.  
\item[7.] Two levels for $\beta_{j1}$, the unstandardized mean difference in treatment outcomes. The absolute value of the treatment-effect difference was varied to be $|\beta_{j1}| = 2.5$ (corresponding to a medium effect size, Cohen's $d = 0.5$; \citeNP{Cohe92}) and $|\beta_{j1}| = 5.0$ (corresponding to a large effect size; Cohen's $d = 1.0$). 
\end{enumerate}

Again, 50 datasets were generated for each cell of the design. As a result, $50 \times 3 \times 2 \times 2 \times 3 \times 3 \times 3 \times 2 = 32,400$ datasets were generated. In every dataset, the outcome variable was calculated as $y_{i} = x_{i}^{\top} \beta_{j} + z_{i}^{\top} b_{m} + \epsilon_{i}$.

@@

Performance of the algorithms was assessed by means of the total number of nodes in estimated GLM and GLMM trees, the accuracy of the resulting trees and predictive accuracy. An accurately recovered tree was defined as a tree with (1) a total of seven nodes (2) the first split involving variable $U_2$ and a value of $30 \pm 5$, (3) the next split on the left involving variable $U_1$ and a value of $17 \pm 5$, and (4) the next split on the right involving variable $U_5$ and a value of $63 \pm 5$. The allowance of $\pm 5$ equals plus or minus half the population standard deviation of the partitioning variable ($\sigma_{U_k}$).

Predictive accuracy of each method was assessed by calculating the correlation between true and predicted treatment-effect differences in every dataset. To prevent overly optimistic estimates of predictive accuracy \cite{HastyTibs09}, predictive accuracy was assessed using test datasets. Test datasets were generated from the same population as training datasets, but test observations were not drawn from the same clusters as the training observations, but from `new' clusters.

To identify the most important predictors of tree size, tree accuracy and predictive accuracy, we used ANOVAs and GLMs with algorithm type and the parameters of the data-generating process as independent variables. First-order interactions between algorithm type and each of the data-generating parameters were also included. The most important predictors were further assessed with graphical displays.

\subsection{Study I: Results}

@@

@@

\begin{figure}[t!]
\centering
<<treesize-xyplot, fig=TRUE, width=9, height=9>>=
aggdata.size <- aggregate(formula=treesize.values ~ treesize.ind + N + sigmab + corUb, FUN=mean, data=treespecs.long)	
print(xyplot(treesize.values ~ sigmab | N + corUb, data = aggdata.size, groups=treesize.ind, 
type="b", ylab="tree size", xlab=expression(sigma[b]), par.settings=standard.theme("pdf",color=F), 
abline=c(7,0), auto.key=list(space="top", columns=2, title=" ", cex.title=1,lines=T, points=T)))
@
\caption{Average tree size of GLM and GLMM trees for datasets with treatment-subgroup interactions. Rows represent levels of dependence between random effects ($b$) and one of the partitioning variables $U_k$; columns represent levels of sample size.}
\label{fig:xyplot_treesize_interact}
\end{figure}

GLMM tree yielded trees with an average size of 7.15 nodes ($\mathrm{SD}=0.61$), whereas GLM tree yielded an trees with an average size of 8.15 nodes ($\mathrm{SD}=2.05$), indicating that GLM trees involved more spurious splits than GLMM trees on average. The estimated probability that a dataset was erroneously not partitioned (the Type-II error) was 0 for both algorithms.

An ANOVA indicated that the most important predictors of tree size were algorithm type, sample size, variance of the random intercept, and the correlation between partitioning variables. The graphical display in Figure~\ref{fig:xyplot_treesize_interact} indicates that GLMM tree grows trees of about the true tree size in all conditions. In the absence of random effects (i.e., $\sigma_{b}=0$), GLM and GLMM trees grow trees of about the same size. However, clear differences in tree size were observed when $\sigma_b > 0$. When sample size is small and the random intercept is not correlated with a partitioning covariate, GLM tree lacks power and grows trees that are too small, on average. When sample size is larger and random intercepts are not correlated with a partitioning covariate, GLM and GLMM trees are about equally sized. However, when random intercepts are correlated with a partitioning covariate, GLM starts to create spurious splits, especially with larger sample sizes and larger $\sigma_b$ values.

@@

Further splits were more accurately recovered by GLMM tree, which accurately recovered the partition in 90.40\% of datasets, and GLM tree in only 61.44\% of datasets. A GLM with tree accuracy as the outcome variable indicates that the most important predictors of tree accuracy were algorithm type, sample size, variance of the random intercept, and level of dependence between the partitioning variables and the random intercept. The graphical display in Figure~\ref{fig:xyplot_treeaccuracy} indicates that in the absence of random effects, the trees recovered by GLM and GLMM tree were about equally accurate. In the presence of random effects, GLM trees were much less accurate than GLMM trees when random effects were correlated with a partitioning covariate. When random intercepts were not correlated with one of the $U_k$ variables, GLMM tree outperformed GLM tree when sample size was small (i.e., $N = 200$).
